---
title: "GSE Report"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{typearea}
        \storeareas\normalpapersize
documentclass: scrreprt
keep-tex: true
---

```{r setup, include = FALSE}
source("survey-data-processing.R")
N_response = nrow(tidy_survey_data)
library(kableExtra)
library(ggpubr)
library(cowplot)
```

```{r base-plot, include=FALSE}
label_truncate = function(string, length){
  str_trunc(string, length, side = "right")
}

base_plot = function(label_length) {
  ggplot(tidy_survey_data) + 
  theme(axis.text.x = element_text(angle=60, hjust=1), legend.position = "bottom") +
  scale_x_discrete(labels = ~label_truncate(., length = label_length)) +
  ylim(0, NA) 
}
```

# Descriptive Statistics

## Respondent Needs

```{r unmet-needs, echo=FALSE}
summarise(tidy_survey_data,
          'Food' = round(sum(NEEDFOOD, na.rm = TRUE)*100/N_response, 1),
          'Medical / Medicine' = round(sum(NEEDMEDICAL, na.rm = TRUE)*100/N_response, 1),
          'Mental Health' = round(sum(NEEDMENTAL, na.rm = TRUE)*100/N_response, 1),
          'Other' = round(sum(NEEDOTHER, na.rm = TRUE)*100/N_response, 1)) %>%
  kableExtra::kable(caption = "\\% Of Respondents Going Without Needs") %>%
  kable_styling(position = "center")
```

```{r rent, echo=FALSE}
base_plot(18) +
  geom_bar(aes(x = PERINCRENT, fill = ROOMIE), position = "stack") +
  labs(x = "Pct. of Income Spent on Rent", y = "Response Count", fill = "Roomate") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey")
```

```{r rent-no-roommate, echo=FALSE}
base_plot(18) +
  geom_bar(aes(x = PERINCTMR)) +
  labs(x = "Pct. of Income Spent on Rent if No Roommate", y = "Response Count")
```

```{r transportation, echo=FALSE, warning=FALSE, message=FALSE}
base_plot(18) +
  geom_bar(aes(x = TRANSPO)) +
  labs(x = "Pct. of Income Spent on Transportation", y = "Response Count") +
  scale_x_discrete(limits = c(levels(tidy_survey_data$PERINCRENT), NA))
  
```

```{r academics, echo=FALSE}
base_plot(18) +
  geom_bar(aes(x = ACADCOST)) +
  labs(x = "Pct. of Income Spent on Academic Costs", y = "Response Count")
```

```{r unexpected-expense, echo=FALSE}
base_plot(18) +
  geom_bar(aes(x = EXPNSE)) +
  labs(x = "Able to Pay Unexpected Emergency Expense", y = "Response Count")
```

```{r netincome, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(tidy_survey_data) +
  geom_histogram(aes(x = NETINC, fill = EXTRNLFUNDED), bins = 20) +
  labs(x = "Monthly Individual Net Income ($)", y = "Response Count", fill = "External Funding") +
  lims(x = c(1000, 4000), y = c(0, 20)) + 
  theme(axis.text.x = element_text(angle=60, hjust=1), legend.position = "bottom") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey")
```

```{r current-sem-support, echo=FALSE}
summarise(tidy_survey_data,
          'Friends, Family, Spouse' = round(sum(PERSONALREL1, na.rm = TRUE)*100/N_response, 1),
          'University Scholarship' = round(sum(UNIVSCHOLSHP1, na.rm = TRUE)*100/N_response, 1),
          'Food Bank' = round(sum(FOODBANK1, na.rm = TRUE)*100/N_response, 1),
          'External Orgs' = round(sum(EXTORG1, na.rm = TRUE)*100/N_response, 1),
          'Loans' = round(sum(LOANS1, na.rm = TRUE)*100/N_response, 1),
          'External Work' = round(sum(EXTWORK1, na.rm = TRUE)*100/N_response, 1)) %>%
  kableExtra::kable(caption = "\\% Of Respondents Relying on Support This Semester") %>%
  kable_styling(position = "center", latex_options = "scale_down")
```

```{r previous-sem-support, echo=FALSE}
summarise(tidy_survey_data,
          'Friends, Family, Spouse' = round(sum(PERSONALREL2, na.rm = TRUE)*100/N_response, 1),
          'University Scholarship' = round(sum(UNIVSCHOLSHP2, na.rm = TRUE)*100/N_response, 1),
          'Food Bank' = round(sum(FOODBANK2, na.rm = TRUE)*100/N_response, 1),
          'External Orgs' = round(sum(EXTORG2, na.rm = TRUE)*100/N_response, 1),
          'Loans' = round(sum(LOANS2, na.rm = TRUE)*100/N_response, 1),
          'External Work' = round(sum(EXTWORK2, na.rm = TRUE)*100/N_response, 1)) %>%
  kableExtra::kable(caption = "\\% Of Respondents Relying on Support in Previous Semesters", format = "latex") %>%
  kable_styling(position = "center", latex_options = "scale_down")
```

```{r work-hrs, echo=FALSE}
base_plot(18) +
  geom_bar(aes(x = HRSWORK)) +
  labs(x = "Hours Worked per Week", y = "Response Count")
```

<!-- Create a landscape page for the big plot -->

```{=tex}
\cleardoublepage
\KOMAoptions{paper=landscape,pagesize}
\recalctypearea
\areaset{1.2\textwidth}{1.2\textheight}
```
## Crosstabs - Years in Program

```{r yip-funding, echo = FALSE}
yip_fund_plot = base_plot(12) +
  geom_bar(aes(x = YEAR, fill = EXTRNLFUNDED), position = "fill") +
  labs(x = "Year in Program", y = "Proportion", fill = "External Funds") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +       
  theme(text = element_text(size = 8))  
```

```{r yip-without, echo = FALSE}
yip_without_plot = base_plot(12) +
  geom_bar(aes(x = YEAR, fill = WITHOUT), position = "fill") +
  labs(x = "Year in Program", y = "Proportion", fill = "Going Without") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 8))  
```

```{r yip-unexpense, echo = FALSE}
yip_unexpense_plot = base_plot(12) +
  geom_bar(aes(x = YEAR, fill = EXPNSE), position = "fill") +
  labs(x = "Year in Program", y = "Proportion", fill = "Handle Unexp. Expn.") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 8))  
```

```{r yip-loans, echo = FALSE}
yip_loans_plot = base_plot(12) +
  geom_bar(aes(x = YEAR, fill = (LOANS1 | LOANS2)), position = "fill") +
  labs(x = "Year in Program", y = "Proportion", fill = "Taken Loans") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 8))  
```

```{r yip-plot-out, echo = FALSE, out.width = '100%', out.height = '100%'}
plot_grid(yip_fund_plot, yip_without_plot, yip_unexpense_plot, yip_loans_plot,
          ncol = 2, nrow = 2)
```

```{r yip-rent, echo = FALSE, out.width = '100%', out.height = '100%'}
base_plot(12) +
  geom_bar(aes(x = YEAR, fill = PERINCRENT), position = "fill") +
  labs(x = "Year in Program", y = "Proportion", fill = "Percent of Income spent on Rent") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 10))  
```

## Crosstabs - Externally Funded

```{r extnl-year, echo = FALSE, out.width = '100%', out.height = '100%'}
extnl_year_plot = base_plot(12) +
  geom_bar(aes(x = EXTRNLFUNDED, fill = YEAR), position = "fill") +
  labs(x = "Externally Funded", y = "Proportion", fill = "Year in Program") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +       
  theme(text = element_text(size = 8))  

extnl_year_plot
```

```{r extnl-without, echo = FALSE}
extnl_without_plot = base_plot(12) +
  geom_bar(aes(x = EXTRNLFUNDED, fill = WITHOUT), position = "stack") +
  labs(x = "Externally Funded", y = "Count", fill = "Going Without") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 8))  
```

```{r extnl-unexpense, echo = FALSE}
extnl_unexpense_plot = base_plot(12) +
  geom_bar(aes(x = EXTRNLFUNDED, fill = EXPNSE), position = "stack") +
  labs(x = "Externally Funded", y = "Count", fill = "Handle Unexp. Expn.") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 8))  
```

```{r extnl-loans, echo = FALSE}
extnl_loans_plot = base_plot(12) +
  geom_bar(aes(x = EXTRNLFUNDED, fill = (LOANS1 | LOANS2)), position = "stack") +
  labs(x = "Externally Funded", y = "Count", fill = "Taken Loans") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 8))  
```

```{r extnl-plot-out, echo = FALSE, out.width = '100%', out.height = '100%'}
plot_grid(extnl_without_plot, extnl_unexpense_plot, extnl_loans_plot,
          ncol = 2, nrow = 2)
```

```{r extnl-rent, echo = FALSE, out.width = '100%', out.height = '100%'}
base_plot(12) +
  geom_bar(aes(x = EXTRNLFUNDED, fill = PERINCRENT), position = "fill") +
  labs(x = "Externally Funded", y = "Proportion", fill = "Percent of Income spent on Rent") +
  scale_fill_viridis_d(direction = -1, na.value="darkgrey") +        
  theme(text = element_text(size = 10))  
```

## Response Demographics

```{r department, echo=FALSE}
dept_plot = base_plot(12) +
  geom_bar(aes(x = fct_infreq(DEPT))) +
  labs(x = "Department", y = "Responses")
```

```{r year, echo=FALSE}
year_plot = base_plot(12) +
  geom_bar(aes(x = YEAR)) +
  labs(x = "Program Year", y = "Responses")
```

```{r age, echo=FALSE}
age_plot = base_plot(12) +
  geom_bar(aes(x = AGE)) +
  labs(x = "Age Bracket", y = "Responses")
```

```{r raceth, echo=FALSE}
raceth_plot = base_plot(12) +
  geom_bar(aes(x = fct_infreq(RACETH))) +
  labs(x = "DARS Ethnicity Category", y = "Responses")
```

```{r demog-plot-out, echo = FALSE, out.width = '100%', out.height = '100%'}
plot_grid(dept_plot, year_plot, age_plot, raceth_plot,
          ncol = 2, nrow = 2)
```

<!-- Return to portrait page -->

```{=tex}
\cleardoublepage
\normalpapersize
\recalctypearea
```
## Budget Example

Suppose a 'typical' student based on these responses (median central
tendencies):

-   They are on a 9-month assistantship contract
-   They are paid the new minimum stipend of \$22,222.22
-   They do not have an external funding source or are not eligible to
    seek external employment

Based on the 2023 fair rent rates [from the Department of Housing and
Urban Development](https://www.huduser.gov/portal/datasets/fmr.html)
which are under the median prices for the area, the student can expect
to pay \$900 monthly for a one bedroom apartment, or \$1,020 monthly for
a two bedroom apartment (\$510 if split with a roommate). The expected
electric bill averaged over the year using Bryan Utility rates is
\$100/mo for the one bedroom, or \$150/mo for the two bedroom (\$75/mo
if split with a roommate). The expected water & wastewater bill is
\$30/mo or \$20/mo with a roommate. The expected monthly internet
expense for B/CS not including cable is \$50 in areas with one provider
and \$40 in areas with more than one provider. Many apartments cover
additional utilities such as trash, but others charge an additional
\$20/mo. In total this works out to \$1070-1100/mo for essential living
expenses living without a roommate or \$625-650/mo with one. Based on
income, this works out to approximately 35% of income with a roommate or
58% of income without a roommate. These figures are consistent with the
observed survey responses. Based on 2019 rates, these costs would be
significantly different; approximately 49% of income without a roommate
and 31% of income with a roommate.[^1]

[^1]: While it is anecdotal; as someone who has lived in Bryan/College
    Station since 2018 these price figures have tracked to my own living
    expenses within about 5% year over year

Based on the survey responses, the median transportation cost is
approximately 15% of income or about \$3,300.

Based on the survey responses, since the majority of people did not
report academic costs it won't be included.

Based on the [December 2022 report from the
USDA](https://www.fns.usda.gov/sites/default/files/media/file/CostofFoodDec2022Thrifty.pdf),
the expected cost of food following a thrifty grocery plan is
\$240-300/mo; or about 13% of income. Based on 2019 rates, these costs
would be significantly different; approximately 9% of income.

With these rates in mind; we can project that an aggressively thrifty
student who relies only on their assistantship for their income is
projected to spend about 63% of their income only on getting to and from
work, a place to live, and food. A thrifty student without a roommate is
projected to spend about 86% on the same.

Notably; these figures do not include other expenses such as insurance,
medical/healthcare, debt payments and saving, or personal spending.
